A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

@@

\begin{lemma}[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
\afterDot
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ necessarily has a
single constructor, whose arguments have types that are interpreted as
singletons.
\end{lemma}

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initially:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

\begin{example}
Let $\const{C} : $ $\tau \times \delta \rightarrow \delta$ be a constructor,
and let $\Ec = \{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x,\, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
Constructing $\Val$ as described above yields
\[\begin{array}{rcrlrl}
\Val \ec{x} & \,=\, &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, & \mu {\VAREC{\const{C}(y, x)}}. & \const{C}( {\Varec{y}},\, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\VAREC{\const{C}(y,x)}}. & \const{C}( {\Varec{y}},\, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, {\VAREC{\const{C}(y,x)}} ) )
\end{array}\]
Since $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are $\alpha$-equivalent,
%for the renaming
%$\{ {\Varec{x}} \mapsto {\VAREC{\const{C}(y,x)}}, {\VAREC{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
the calculus can infer $x \teq \const{C}(y,x)$.
%Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
%are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t : \tau \in \tEc
  \quad
  ( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\tau )
  \text{ or }
  ( \tau \in \Data \text{ and } \tau \text{ finite} )
}{
  \Ec := \Ec,\: t \teq \const{C}_1(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec,\: t \teq \const{C}_m(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) )
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Phase 3: Branching.}
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type,
the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$).
This is enforced in the third phase by the \rn{Split} rule.
The \rn{Singleton} rule handles the degenerate case where two terms from $\tEc$
have a corecursive singleton type, in which case they must be equivalent.

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu {\Varec{y_1}}. & \const{A}( & \mu {\Varec{y_2}}. & \const{B}( {\Varec{x}} ) ) &
\Val \ec{w_1} & = & \mu {\Varec{w_1}}. & \const{A}( {\Varec{x}} )      & \\
\Val \ec{y_2} & = & \mu {\Varec{y_2}}. & \multicolumn{2}{l}{\const{B}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \mu {\Varec{w_2}}. & \const{C}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \mu {\Varec{z_1}}. & \const{C}( & \mu {\Varec{z_2}}. & \const{A}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \mu {\Varec{z_2}}. & \const{A}( & \mu {\Varec{z_1}}. & \const{C}( {\Varec{z_2}} ) ) &
\Val \ec{x} & = & {\Varec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$,
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) ) )$,
which has normal form
$\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( {\Varec{w_1}} ) )$.
\xend
\end{example}

>>>>

\makeatletter
\def\spnXwtheorem#1#2#3#4{\@spynthm{#1}{#2}{#3}{#4}%
                         \@addtoreset{#1}{chapter}}%
\makeatother

%%% not part of shared counter
\spnXwtheorem{examplex}{Example}{\itshape}{\rmfamily}

@@

\begin{lemma}%[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
\afterDot
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ necessarily has a
single constructor, whose arguments have types that are interpreted as
singletons.
\end{lemma}

@@

\begin{examplex}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Unfold} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initially:}\phantom{Unfold $\Varec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{y}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Unfold $\Varec{x}$:}\phantom{Unfold $\Varec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

\begin{examplex}
Let $\const{C} : $ $\tau \times \delta \rightarrow \delta$ be a constructor,
and let $\Ec = \{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x,\, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
Constructing $\Val$ as described above yields
\[\begin{array}{rcrlrl}
\Val \ec{x} & \,=\, &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, & \smash{\mu {\vthinspace\VAREC{\const{C}(y, x)}}}. & \const{C}( {\Varec{y}},\, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\vthinspace\VAREC{\const{C}(y,x)}}. & \const{C}( {\Varec{y}},\, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}},\, {\VAREC{\const{C}(y,x)}} ) )
\end{array}\]
Since $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are $\alpha$-equivalent,
%for the renaming
%$\{ {\Varec{x}} \mapsto {\VAREC{\const{C}(y,x)}}, {\VAREC{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
the calculus can infer $x \teq \const{C}(y,x)$.
%Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
%are equivalent for any value of ${\Varec{y}}$.
\xend
\end{examplex}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t : \delta \in \tEc
  \quad
  ( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta )
  \text{ or }
  ( \delta \in \Data \text{ and } \delta \text{ finite} )
}{
  \Ec := \Ec,\: t \teq \const{C}_1(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec,\: t \teq \const{C}_m(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) )
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  \delta \in \Codata
  \quad
  \left|\delta\right| = 1
  \quad
  t, u : \delta \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain an application of one of the constructors of $\delta$ (where $\Ctr^\delta = \{ \const{C}_1, \ldots \const{C}_m \}$).
This is enforced in the third phase by the \rn{Split} rule.
The \rn{Singleton} rule handles the degenerate case where two terms from $\tEc$
have a corecursive singleton type, in which case they must be equivalent.

@@

\begin{examplex}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu {\Varec{y_1}}. & \const{A}( & \mu {\Varec{y_2}}. & \const{B}( {\Varec{x}} ) ) &
\Val \ec{w_1} & = & \mu {\Varec{w_1}}. & \const{A}( {\Varec{x}} )      & \\
\Val \ec{y_2} & = & \mu {\Varec{y_2}}. & \multicolumn{2}{l}{\const{B}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \mu {\Varec{w_2}}. & \const{C}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \mu {\Varec{z_1}}. & \const{C}( & \mu {\Varec{z_2}}. & \const{A}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \mu {\Varec{z_2}}. & \const{A}( & \mu {\Varec{z_1}}. & \const{C}( {\Varec{z_2}} ) ) &
\Val \ec{x} & = & {\Varec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$,
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) ) )$,
which has normal form
$\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}
